The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2014-12-04 draft.tex
Some more CK suggested changes, added refs, option to add Ly alpha figure

It has been known for a long time that winds emanating from the
accretion disk are important in shaping the ultraviolet (UV) spectra
of high-state CVs (Heap 1978). The most spectacular evidence for such
outflows are the P-Cygni-like profiles seen in UV resonance lines such as
\civfull\ (see e.g. Cordova \& Mason
1982\nocite{cordova1982}). Considerable effort has been spent over the
years on understanding and modelling these UV features (e.g. Drew \&
Verbunt 1985; Mauche \& Raymond
1987; Drew 1987; Shlosman \& Vitello 1993; [hereafter
SV93]\nocite{SV93}; Knigge, Woods \& Drew 1995 [hereafter
KWD95]\nocite{KWD95}; Knigge \& Drew 1997; Knigge et al. 1997; Long \&
Knigge 2002 [hereafter LK02]\nocite{LK02}, Noebauer et al. 2010;
Puebla et al. 2011). The basic picture emerging from these efforts is
of a slowly accelerating, moderately collimated bipolar
outflow that carries away $\simeq 1\% - 10\%$ of the accreting
material. State-of-art-simulations of line formation in this type
of disk wind can produce UV line profiles that are remarkably similar
to observations.

@@

Murray \& Chiang (1996; hereafter MC96)\nocite{MC96} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse (REF). This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{RN98}.
{\bf JM: need to mention this as pointed out by CK.}

@@

Attempts to fit the observed spectral energy distributions (SEDs) of
high-state CVs with such models have met with mixed success. In
particular, the SEDs predicted by most stellar/disk atmosphere models 
are too blue in the UV (REFs) and exhibit
stronger-than-observed Balmer jumps in absorption (REFs). One possible
explanation for these problems is that these models fail to capture
all of the relevant physics. Indeed, it has been argued that a
self-consistent treatment can 
produce better agreement with observational data (e.g. Shaviv et
al. 1991;  but also more recent work by Idan et al. 2010). 
\nocite{idanshaviv2010} \nocite{shaviv1991}. However, an alternative 
explanation, suggested by \cite{KLWB98}, is that recombination continuum
emission from the base of the disk wind might fill in the disk's
Balmer absorption edge and flatten the UV spectrum.

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig1.eps}
\caption{Optical spectra of three nova-like variables: 
RW Sex (Top; Beuermann et al. 1992),
IX Vel (Middle; Beuermann \& Thomas 1990) %% CITE THESE
and RW Tri in eclipse (Bottom; Groot et al. 2004).
The data have been digitized from the respective publications and a few of the more
prominent emission and absorption lines are marked. 
These systems have inclinations of $30^\circ$, $60^\circ$ and $70^\circ$ 
respectively.
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum.}
\label{novalikes}
\end{figure}

%% CITATIONS FROM CAPTION
\nocite{groot2004}

@@

Our main goal is to test whether the type of disk wind model that has
been successful in explaining the UV spectra of CVs could also have 
significant impact on the optical continuum and emission line spectra
of these systems. In order to set a benchmark, we therefore begin by
investigating one of the fiducial CV wind models that was used by SV93
and LK02 to simulate the UV spectrum of a typical high-state
system. The specific parameters for this model are listed in
Table~1. A key point is that the wind mass-loss rate in this model is
set to 10$\%$ of the accretion rate through the disk.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/3by3_grid_balmer.eps}  
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
The evolution of the Balmer jump with varying kinematic 
properties, computed for an inclination of $80^\circ$. 
The panels are organised in the same way as figure~\ref{halpha}.
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{jump}
\end{figure} %fullpage

@@

\begin{figure} %fullpage
\mbox{
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_uv.eps}}
\quad
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_1640.eps}}   
}
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
Line profiles for \civ\ (left) and \heiiuv\  (right) with varying kinematic 
properties, computed for an inclination of $80^\circ$. 
The panels are organised in the same way as figure~\ref{halpha}.
{\bf JM: I need to adjust the smoothing here so for all plots I smooth
in terms of Wavelength rather than bins. Linked to single-peaked discussion. 
Note also CK may want a Lyman alpha plot, which I might put here}
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{uvlines}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{p{3cm}p{4cm}}
Revised Model (Optimized for the Optical Band)\\
\hline Free Parameters 	&	 Value \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$ \\ 
$\dot{M}_{acc}$ 	 &	 $10^{-9}~M_{\odot}yr^{-1}$\\ 
$\dot{M}_{wind}$  &	$10^{-8}~M_{\odot}yr^{-1}$\\ 
$r_{min}$ 	&	 $4 R_{WD}$\\ 
$r_{max}$ 	&	 $12 R_{WD}$ \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$ \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$ \\ 
$\gamma$ 	&	 $1$ \\ 
$v_{\infty}$ 	&	 $3v_{esc}$ \\ 
$R_v$ 	        &	 $142.9 R_{WD}$ \\ 
$\alpha$ 	&	 $4$ \\
\end{tabular}
\centering
\caption{Wind geometry parameters used in the next-generation CV
  model. {\bf CK: I actually think we should probably have a single
    table with both sets of model pars}}
\label{modelb_table}
\end{table}

The benchmark model discussed in section~\ref{modela} was originally
designed to reproduce the wind-formed lines seen in the UV spectra of
high-state CVs. As we have seen, this model does produce significant
amounts of optical emission. However, with only relatively minor
changes to the kinematics of the benchmark model, it is possible to
construct a model that more closely matches the observed {\em optical}
spectra of CVs. Guided by the results of the previous section, we
adopt the parameters listed in table~\ref{modelb_table} for this purpose.

>>>>

It has been known for a long time that winds emanating from the
accretion disk are important in shaping the ultraviolet (UV) spectra
of high-state CVs (Heap 1978). The most spectacular evidence for such
outflows are the P-Cygni-like profiles seen in UV resonance lines such as
\civfull\ (see e.g. Cordova \& Mason
1982\nocite{cordova1982}). Considerable effort has been spent over the
years on understanding and modelling these UV features (e.g. Drew \&
Verbunt 1985\nocite{drewverbunt1985}; Mauche \& Raymond
1987\nocite{maucheraymond1987}; Drew 1987; Shlosman \& Vitello 1993; [hereafter
SV93]\nocite{SV93}; Knigge, Woods \& Drew 1995 [hereafter
KWD95]\nocite{KWD95}; Knigge \& Drew 1997\nocite{kd1997}; 
Knigge et al. 1997\nocite{knigge1997}; Long \&
Knigge 2002 [hereafter LK02]\nocite{LK02}, Noebauer et al. 2010;
Puebla et al. 2011\nocite{puebla2011}). The basic picture emerging from these efforts is
of a slowly accelerating, moderately collimated bipolar
outflow that carries away $\simeq 1\% - 10\%$ of the accreting
material. State-of-art-simulations of line formation in this type
of disk wind can produce UV line profiles that are remarkably similar
to observations.

@@

Murray \& Chiang (1996; hereafter MC96)\nocite{MC96} 
have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse (REF). This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{RN98, kafka2004}.
{\bf JM: need to mention this later as pointed out by CK- both in discussion
and conclusion sections.}

@@

Attempts to fit the observed spectral energy distributions (SEDs) of
high-state CVs with such models have met with mixed success. In
particular, the SEDs predicted by most stellar/disk atmosphere models 
are too blue in the UV \citep{wade1988,long1991,long1994}(Other refs here?) and exhibit
stronger-than-observed Balmer jumps in absorption \cite{wade1984,haug1987,ladous1989b,knigge1998}(Other refs here?). One possible
explanation for these problems is that these models fail to capture
all of the relevant physics. Indeed, it has been argued that a
self-consistent treatment can 
produce better agreement with observational data (e.g. Shaviv et
al. 1991;  but also more recent work by Idan et al. 2010). 
\nocite{idanshaviv2010} \nocite{shaviv1991}. However, an alternative 
explanation, suggested by \cite{KLWB98}, is that recombination continuum
emission from the base of the disk wind might fill in the disk's
Balmer absorption edge and flatten the UV spectrum.

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig1.eps}
\caption{Optical spectra of three nova-like variables: 
RW Sex (top; Beuermann et al. 1992),
IX Vel (top middle; Beuermann \& Thomas 1990) 
and RW Tri in and out of eclipse (bottom two panels; Groot et al. 2004).
The data have been digitized from the respective publications and a few of the more
prominent emission and absorption lines are marked. 
These systems have inclinations of $30^\circ$, $60^\circ$ and $70^\circ$ 
respectively.
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum.}
\label{novalikes}
\end{figure}

%% CITATIONS FROM CAPTION
\nocite{groot2004}
\nocite{beuermann1990}
\nocite{beuermann1992}

@@

Our main goal is to test whether the type of disk wind model that has
been successful in explaining the UV spectra of CVs could also have 
significant impact on the optical continuum and emission line spectra
of these systems. In order to set a benchmark, we therefore begin by
investigating one of the fiducial CV wind models that was used by SV93
and LK02 to simulate the UV spectrum of a typical high-state
system. The specific parameters for this model (Model A) are listed in
Table~1. A key point is that the wind mass-loss rate in this model is
set to 10$\%$ of the accretion rate through the disk.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/3by3_grid_balmer.eps}  
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
The evolution of the Balmer jump with varying kinematic 
properties, computed for an inclination of $80^\circ$. 
The panels are organised in the same way as figure~\ref{halpha}.
{\bf JM: Mark lines here?}
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{jump}
\end{figure} %fullpage

@@

\begin{figure} %fullpage
\mbox{
% \subfigure{\includegraphics[width=0.33\textwidth]{figures/3by3_grid_lyman.eps}}  
% \quad
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_uv.eps}}
\quad
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_1640.eps}}  
}
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
Line profiles for \civ\ (left) and \heiiuv\  (right) with varying kinematic 
properties, computed for an inclination of $80^\circ$. 
The panels are organised in the same way as figure~\ref{halpha}.
{\bf JM: I need to adjust the smoothing here so for all plots I smooth
in terms of Wavelength rather than bins. Linked to single-peaked discussion. 
Note also CK may want a Lyman alpha plot, which I might put here}
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{uvlines}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{p{3cm}p{4cm}}
Model B\\
\hline Free Parameters 	&	 Value \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$ \\ 
$\dot{M}_{acc}$ 	 &	 $10^{-9}~M_{\odot}yr^{-1}$\\ 
$\dot{M}_{wind}$  &	$10^{-8}~M_{\odot}yr^{-1}$\\ 
$r_{min}$ 	&	 $4 R_{WD}$\\ 
$r_{max}$ 	&	 $12 R_{WD}$ \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$ \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$ \\ 
$\gamma$ 	&	 $1$ \\ 
$v_{\infty}$ 	&	 $3v_{esc}$ \\ 
$R_v$ 	        &	 $142.9 R_{WD}$ \\ 
$\alpha$ 	&	 $4$ \\
\end{tabular}
\centering
\caption{Wind geometry parameters used in the revised CV model, Model B,
which is {\em optimized for the optical band}
{\bf CK: I actually think we should probably have a single
table with both sets of model pars}}
\label{modelb_table}
\end{table}

The benchmark model discussed in section~\ref{modela} was originally
designed to reproduce the wind-formed lines seen in the UV spectra of
high-state CVs. As we have seen, this model does produce significant
amounts of optical emission. However, with only relatively minor
changes to the kinematics of the benchmark model, it is possible to
construct a model that more closely matches the observed {\em optical}
spectra of CVs (imaginatively dubbed `Model B'). Guided by the results of the previous section, we
adopt the parameters listed in table~\ref{modelb_table} for this purpose.
